### 2022-07-25 paper/localpsf.tex
more paper edits. Barely under 25 pages

\begin{figure}[H]
	\begin{subfigure}{0.5\textwidth}
		\includegraphics[scale=0.23]{meshHeight_view2_edges.png}
		\caption{Bird's eye view}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\includegraphics[scale=0.22]{meshHeight_view0_edges_withAxes.png}
		\caption{Top-down view}
	\end{subfigure}
	\caption{The ice sheet, discretized by a mesh of tetrahedra. Color indicates the height of the base of the ice sheet (i.e., the mountain topography). The radius of the domain is $10$ kilometers, and the maximum height of the mountain is $2.1$ kilometers.}
	\label{fig:ice_mountain_mesh}
\end{figure}

@@

The reconstructed parameter field $\basalfriction$ is found as the solution to an optimization problem of the following form:
\begin{equation}
	\label{eq:deterministic_optimization_problem}
	\min_{\basalfriction} \quad J(\basalfriction) := J_d(\basalfriction) + J_r(\basalfriction).
\end{equation}
The first term in the objective function is
\begin{equation*}
	J_d(\basalfriction) := \frac{1}{2}\int_{\Gamma_{t}}||\observations_\text{obs} - \tangentop\velocity||_{2}^{2}\,dS.
\end{equation*}
We call this term the data misfit, because it measures the mismatch between the observed data and the predicted data based on a candidate parameter $\basalfriction$. Here $\velocity = \velocity(\basalfriction)$ denotes the velocity field solving $\ref{Stokeseqn}$ for the given parameter $\basalfriction$. The second term,
\begin{equation*}
	J_r(\basalfriction) := \frac{1}{2}\int_{\Omega}\left| \mathcal{L}(\basalfriction - \basalfriction_0)\right|^2\,dx,
\end{equation*}
is a Bilaplacian regularization term. Specifically, $\mathcal{L}$ is the inverse of the solution operator for the following elliptic PDE:
\begin{align*}
-\gamma\Delta u +\delta\,u&=s\,\,\,\,\text{ in }\Omega,\\
\gamma \nabla u\cdot\normalvec+\mu\,u&=0\,\,\,\text{ on }\partial\Omega.
\end{align*}
Recall that $\Omega$ is the 2D flattened version of the basal surface $\Gamma_b$.
In all results shown except one, $\gamma$ is chosen so that the solution satisfies the Morozov discrepancy principle, that is,
\begin{equation*}
	\frac{1}{2}\int_{\Gamma_{t}}||\observations_\text{obs} - \tangentop\velocity||_{2}^{2}\,dS = \frac{1}{2}\int_{\Gamma_{t}}||\xi||_{2}^{2}\,dS.
\end{equation*}
The value of $\gamma$ satisfying the Morozov discrepancy principle is $\gamma=7.3 \times 10^3$. The exception is Figure \ref{fig:gamma_sweep}, where we vary $\gamma$ to study the effect of changing $\gamma$ on the quality of our preconditioner. The constant $\delta$ is chosen so that the correlation length, $L$, of functions drawn from the normal distribution with covariance $\mathcal{L}^{-2}$, which is given by the formula $L=\sqrt{\gamma/\delta}$, is $1/10$th of the radius of the domain. The function $\basalfriction_0$ is the constant function with value $\basalfriction_0(x)=10.5$, which is halfway between the maximum and minimum values of $\basalfriction_\text{true}$.

The inverse problem is solved with a inexact Newton preconditioned conjugate gradient scheme (PCG) \cite{NOCEDAWRIGHT}. The constant function $\basalfriction_0$ is used as the initial guess. The parameter value is updated iteratively according to the formula
$
	\basalfriction_{k+1} = \basalfriction_k + \alpha_k \searchdir_k,
$
where $\searchdir_k$ approximately solves one of the following Newton linear systems
\begin{equation}
	\label{eq:newton_system}
	H \searchdir_k = - g_k \qquad \text{or} \qquad H_\text{gn}, \searchdir_k = - g_k
\end{equation}
and the step length $\alpha_k \in \mathbb{R}$ is chosen to satisfy the Armijo conditions \cite{ARMIJO}. Here $g_k$ is the gradient of $J$, $H$ is the Hessian of $J$, and $H_\text{gn}$ is the Gauss-Newton Hessian, and these derivatives are evaluated at $\basalfriction_k$. 
%The Gauss-Newton Hessian is Hessian of $J$, except in the data misfit, $J_d$, the nonlinear mapping $\basalfriction \mapsto \tangentop \velocity$ is replaced by its local linearization at $\basalfriction_k$.
We use $H_\text{gn}$ for the first five iterations, and $H$ for all subsequent iterations.  The method terminates when $||g_k||/||g_0|| < 10^{-8}$. The systems \eqref{eq:newton_system} are solved inexactly using an inner PCG iteration, which terminates when the norm of the residual for the linear system is less than $\min \left(0.5, \sqrt{||g_k||}\right)||g_k||$. The PCG method only requires applying $H$ (or $H_\text{gn}$) to vectors. Each application of $H$ (or $H_\text{gn}$) to a vector requires solving one incremental forward and one incremental adjoint Stokes PDE of the form \eqref{Stokeseqn}, but with different right hand sides.

@@

We use the framework described in this paper to generate Hessian preconditioners. We build H-matrix approximations of the data-misfit Gauss-Newton Hessian (the term in $H_\text{gn}$ that arises from the data misfit) using 1, 5, and 25 impulse responses batches. The resulting H-matrices, which may be indefinite due to approximation error, are modified to be symmetric positive definite via the procedure described in Appendix \ref{sec:make_spd}, with $l=2$. We approximate $H_\text{gn}$ rather than $H$ because $H$ often has negative values in its integral kernel when one is away from the optimal point. Our results show that our preconditioners are also good preconditioners for $H$. The H-matrix approximation of data misfit Gauss-Newton Hessian is added to a H-matrix approximation of the Hessian of the regularization term, $J_r$, to form a H-matrix approximation of the Gauss-Newton Hessian. For the Hessian of the regularization term, straightforward calculation shows that the discrete regularization term in both the Hessian and the Gauss-Newton Hessian is given by
$
(\gamma \bm{K} + \delta \bm{M}) \bm{M}^{-1} (\gamma \bm{K} + \delta \bm{M})
$
where $\bm{K}$ is a finite element stiffness matrix with appropriate boundary conditions, and $\bm{M}$ is the finite element mass matrix. These matrices are sparse, so we form H-matrix representations of these matrices using standard sparse H-matrix techniques. Then we invert $\bm{M}^{-1}$, and multiply these matrices together using H-matrix methods. We factor the overall Gauss-Newton Hessian approximation, denoted $\widetilde{H}$, using fast H-matrix methods, then and use the factorization as a preconditioner \cite{HLIBPRO}. The approximations for 1, 5, and 25 impulse response batches are denoted by PSF (1), PSF (5), and PSF (25), respectively.

@@

In Table \ref{tab:newton_convergence_table} we study the performance of our preconditioner for accelerating the solution of optimization problem \eqref{eq:deterministic_optimization_problem}. We build the PSF (5) preconditioner in the third Gauss-Newton iteration, and re-use it for all subsequent Gauss-Newton and Newton iterations. No preconditioning is used in the three Gauss-Newton iterations before the PSF (5) preconditioner is built. We compare our method with the most commonly used existing preconditioners, which are no preconditioning (denoted by NONE), and preconditioning by the regularization term in the Hessian (denoted REG). In Figure \ref{fig:newton_iterates} we show select Newton iterates using our PSF (5) preconditioner. Using PSF (5) reduces the total number of Stokes PDE solves by a factor of six compared to no preconditioning, and by a factor of nine compared to regularization preconditioning. Interestingly, we observe that no preconditioning actually outperforms regularization preconditioning here because the noise level, and hence the regularization parameter, $\gamma$, are small.

>>>>

\begin{figure}[H]
	\begin{subfigure}{0.5\textwidth}
		\includegraphics[scale=0.23]{meshHeight_view2_edges.png}
		\caption{Bird's eye view}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\includegraphics[scale=0.22]{meshHeight_view0_edges_withAxes.png}
		\caption{Top-down view}
	\end{subfigure}
	\caption{The ice sheet, discretized by a mesh of tetrahedra. Color indicates the height of the base of the ice sheet (i.e., the mountain topography). The radius of the domain is $10$ kilometers, the maximum height of the mountain is $2.1$ kilometers, and the average thickness of the ice sheet is 250 meters.}
	\label{fig:ice_mountain_mesh}
\end{figure}

@@

The reconstructed parameter field $\basalfriction$ is found as the solution to an optimization problem of the following form:
\begin{equation}
	\label{eq:deterministic_optimization_problem}
	\min_{\basalfriction} \quad J(\basalfriction) := J_d(\basalfriction) + J_r(\basalfriction).
\end{equation}
The first term in the objective function is
\begin{equation*}
	J_d(\basalfriction) := \frac{1}{2}\int_{\Gamma_{t}}||\observations_\text{obs} - \tangentop\velocity||_{2}^{2}\,dS.
\end{equation*}
We call this term the data misfit, because it measures the mismatch between the observed data and the predicted data based on a candidate parameter $\basalfriction$. Here $\velocity = \velocity(\basalfriction)$ denotes the velocity field solving $\ref{Stokeseqn}$ for the given parameter $\basalfriction$. The second term,
\begin{equation*}
	J_r(\basalfriction) := \frac{1}{2}\int_{\Omega}\left| \mathcal{L}(\basalfriction - \basalfriction_0)\right|^2\,dx,
\end{equation*}
is a Bilaplacian regularization term. Specifically, $\mathcal{L}$ is the inverse of the solution operator for the following elliptic PDE:
\begin{align*}
-\gamma\Delta u +\delta\,u&=s\,\,\,\,\text{ in }\Omega,\\
\gamma \nabla u\cdot\normalvec+\mu\,u&=0\,\,\,\text{ on }\partial\Omega.
\end{align*}
Recall that $\Omega$ is the 2D flattened version of the basal surface $\Gamma_b$.
In all results shown except one, $\gamma$ is chosen so that the solution satisfies the Morozov discrepancy principle, that is,
\begin{equation*}
	\frac{1}{2}\int_{\Gamma_{t}}||\observations_\text{obs} - \tangentop\velocity||_{2}^{2}\,dS = \frac{1}{2}\int_{\Gamma_{t}}||\xi||_{2}^{2}\,dS.
\end{equation*}
The value of $\gamma$ satisfying the Morozov discrepancy principle is $\gamma=7.3 \times 10^3$. The exception is Figure \ref{fig:gamma_sweep}, where we vary $\gamma$ to study the effect of changing $\gamma$ on the quality of our preconditioner. The constant $\delta$ is chosen so that the correlation length, $L=\sqrt{\gamma/\delta}$, of functions drawn from the normal distribution with covariance $\mathcal{L}^{-2}$, is $1/10$th of the radius of the domain. The function $\basalfriction_0$ is the constant function with value $\basalfriction_0(x)=10.5$.
%, which is halfway between the maximum and minimum values of $\basalfriction_\text{true}$.

The inverse problem is solved with a inexact Newton preconditioned conjugate gradient (PCG) scheme \cite{NOCEDAWRIGHT}. The constant function $\basalfriction_0$ is used as the initial guess. The parameter value is updated iteratively according to the formula
$
	\basalfriction_{k+1} = \basalfriction_k + \alpha_k \searchdir_k,
$
where $\searchdir_k$ approximately solves one of the following Newton linear systems
\begin{equation}
	\label{eq:newton_system}
	H \searchdir_k = - g_k \qquad \text{or} \qquad H_\text{gn}, \searchdir_k = - g_k
\end{equation}
and the step length $\alpha_k \in \mathbb{R}$ is chosen to satisfy the Armijo conditions \cite{ARMIJO}. Here $g_k$ is the gradient of $J$, $H$ is the Hessian of $J$, and $H_\text{gn}$ is the Gauss-Newton Hessian, and these derivatives are evaluated at $\basalfriction_k$. 
%The Gauss-Newton Hessian is Hessian of $J$, except in the data misfit, $J_d$, the nonlinear mapping $\basalfriction \mapsto \tangentop \velocity$ is replaced by its local linearization at $\basalfriction_k$.
We use $H_\text{gn}$ for the first five iterations, and $H$ for all subsequent iterations.  The method terminates when $||g_k||/||g_0|| < 10^{-8}$. The systems \eqref{eq:newton_system} are solved inexactly using an inner PCG iteration, which terminates when the norm of the residual for the linear system is less than $\min \left(0.5, \sqrt{||g_k||}\right)||g_k||$. The PCG method only requires applying $H$ (or $H_\text{gn}$) to vectors. Each application of $H$ (or $H_\text{gn}$) to a vector requires solving one incremental forward and one incremental adjoint Stokes PDE of the form \eqref{Stokeseqn}, but with different right hand sides.

@@

We use the framework described in this paper to generate Hessian preconditioners. We build H-matrix approximations of the data-misfit Gauss-Newton Hessian (the term in $H_\text{gn}$ that arises from the data misfit) using 1, 5, and 25 impulse responses batches. The resulting H-matrices, which may be indefinite due to approximation error, are modified to be symmetric positive definite via the procedure described in Appendix \ref{sec:make_spd}, with $l=2$. We approximate $H_\text{gn}$ rather than $H$ because $H$ often has negative values in its integral kernel when one is away from the optimal point. Our results show that our preconditioners are also good preconditioners for $H$. The H-matrix approximation of data misfit Gauss-Newton Hessian is added to a H-matrix approximation of the Hessian of the regularization term to form a H-matrix approximation of $H_\text{gn}$. Straightforward calculation shows that the discretized Hessian of the regularization term is given by
$
(\gamma \bm{K} + \delta \bm{M}) \bm{M}^{-1} (\gamma \bm{K} + \delta \bm{M})
$
where $\bm{K}$ is a finite element stiffness matrix with appropriate boundary conditions, and $\bm{M}$ is the finite element mass matrix. These matrices are sparse, so we form H-matrix representations of these matrices using standard sparse H-matrix techniques. Then we invert $\bm{M}^{-1}$, and multiply these matrices together using H-matrix methods. We factor the overall Gauss-Newton Hessian approximation, denoted $\widetilde{H}$, using fast H-matrix methods, then and use the factorization as a preconditioner \cite{HLIBPRO}. The approximations for 1, 5, and 25 impulse response batches are denoted by PSF (1), PSF (5), and PSF (25), respectively.

@@

In Table \ref{tab:newton_convergence_table} we study the performance of our preconditioner for accelerating the solution of optimization problem \eqref{eq:deterministic_optimization_problem}. We build the PSF (5) preconditioner in the third Gauss-Newton iteration, and re-use it for all subsequent Gauss-Newton and Newton iterations. No preconditioning is used in the three Gauss-Newton iterations before the PSF (5) preconditioner is built. We compare our method with the most commonly used existing preconditioners, which are no preconditioning (denoted by NONE), and preconditioning by the regularization term in the Hessian (denoted REG). In Figure \ref{fig:newton_iterates} we show select Newton iterates using our PSF (5) preconditioner. Using PSF (5) reduces the total number of Stokes PDE solves by a factor of six compared to no preconditioning, and by a factor of nine compared to regularization preconditioning. Interestingly, we observe that no preconditioning actually outperforms regularization preconditioning here because the noise level (and hence $\gamma$) is small.
